---
title: "Unpaid Carers during the pandemic"
author: ""
date: "5/4/2021"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: paper
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning=FALSE)
library(tidyverse)
library(THFstyle)
library(ggplot2)
library(ggfittext)
library(scales)
library(janitor)
library(networkD3)
library(stringr)
library(survey)
library(ggtext)
library(plotly)
```

```{r}
#Load Data 
all_h<- readRDS(here::here('data','care_type','health_all.rds'))
all_dem<- readRDS(here::here('data','care_type','demographics_all.rds'))
all<- readRDS(here::here('data','care_type','caring_pandemic_care_type.rds'))
all_s<- readRDS(here::here('data','care_type','services_all.rds'))

```
## Section 1 - Background and Methods {.tabset}

### Background
The impact of unpaid caring is closely related to the intensity of the care provision.[1,2,3] Carers within the same household are more likely to provide longer hours of care and more personal care task such as dressing or bathing. Longer hours and personal care responsibilities are associated with poorer psychological and physical health.[1,2,3,4,5,6] This could be due to the notion of not being able to ‘switch off’ the caring duties and have no time for respite or to manage their own health. It is important to focus on carers with high level of care, as they are likely to need further support. 

During the COVID-19 pandemic, carers reported disruptions in their caring responsibilities with some reporting increased care but reduced support.[7] Many reported lower wellbeing, financial difficulties (relying on foodbanks) and lack of respite opportunities. [7,8,9,10] Those that did have access to formal care, were worried about infection control but many reported cancellation or postponement of NHS procedures/treatments for themselves and the person they are caring for. [7,8]. Organisations such as Future Care Capital and Carers UK recognised the need to value their contributions and better support them going forward, particularly with their own health needs. [7,11]. 

However, as prior studies have stipulated, it is important to make a distinction between different caring intensity to fully capture the complex impact of caring. Without such distinction, the needs of carers with high care level (based on caring intensity) may not be fully captured. This analysis explores who the unpaid carers are during the second half of the pandemic, their state of health and how they have managed to access services to manage their own health. It will make a distinction between different caring types to highlight the carers who may benefit from further support.


### Methods

Source: Understanding Society Data set, Wave 10 (2018/19) and COVID-19 sub study Waves 6 (November 2020) and 7 (January 2021)

For the COVID-19 data, wave 6 and wave 7 data set was linked but for individuals who participated in both waves, the most available data was retained. 

Unpaid carers during the pandemic were identified if they answered yes to one of the following:

1. Is there anyone living with you who is sick, disabled or elderly whom you look after or give special help to (for example, a sick, disabled or elderly relative, husband, wife or friend etc)?
2. Thinking about the last 4 weeks, did you provide help or support to family, friends or neighbours who do not live in the same house/flat as you?

Unpaid carers pre pandemic were identified if they answered yes to one of the following:

1. Do you provide regular service or hep for any sick, disabled or elderly person not living with you?” 
2. “Is there anyone living with you who is sick, disabled, or elderly, who you give special help to?”

Level of care is defined as follows:

* High caring level: Carers who  who reported >= 20 hours per week or helped with basic personal tasks. 
* Low caring level: Carers who reported <20 hours of care per week and not helped with basic personal tasks.

Weighted estimates:

All estimates provided have accounted for non-response, sampling probabilities and the sampling design of the Understanding Society Survey. We used the cross-sectional weights provided unless when comparing pre-pandemic, we used the longitudinal weights provided. 

### Code

All code can be found on [github](https://github.com/HFAnalyticsLab/Unpaid_Carers)

### References

[1] Future Care Capital. A forgotten army, coping as a carer. Published Online First 2019.
[https://futurecarecapital.org.uk/latest/joel-charles-unpaid-carers-are-the-backbone-of-our-society-their-priorities-should-be-our-priorities/](https://futurecarecapital.org.uk/latest/joel-charles-unpaid-carers-are-the-backbone-of-our-society-their-priorities-should-be-our-priorities/)

[2] Bom J, Bakx P, Schut F, van Doorslaer E. 2019. The Impact of Informal Caregiving for Older Adults on Health of various types of caregivers: A systematic Review. The Gerontologist 59(5). [https://doi.org/10.1093/geront/gny137](https://doi.org/10.1093/geront/gny137)

[3] Bom J, Stöckel J. Is the grass greener on the other side? The health impact of providing informal care in the UK and the Netherlands. 2021. Social Science & Medicine 269. [https://doi.org/10.1016/j.socscimed.2020.113562](https://doi.org/10.1016/j.socscimed.2020.113562) 

[4] Lacey R, McMunn A, Webb E. Informal caregiving and mental health. 2018. Psychological Medicine 49 (10)[https://doi:10.1017/S0033291718002222](https://doi:10.1017/S0033291718002222)

[5] EuroCarers. The Impact of Caregiving on Informal carers' mental and physical health.[https://eurocarers.org/wp-content/uploads/2018/09/Eurocarers-mental-and-physical-health_final.pdf] (https://eurocarers.org/wp-content/uploads/2018/09/Eurocarers-mental-and-physical-health_final.pdf)

[6] Carers UK. Caring behind closed doors: six months on. Published Online First: 2020.[http://www.carersuk.org/images/News_and_campaigns/Behind_Closed_Doors_2020/Caring_behind_closed_doors_Oct20.pdf](http://www.carersuk.org/images/News_and_campaigns/Behind_Closed_Doors_2020/Caring_behind_closed_doors_Oct20.pdf)

[7] BENNETT MR, ZHANG Y, YEANDLE S. Caring and COVID-19: hunger and mental wellbeing. Care Matters Ser 2020;:16.[http://circle.group.shef.ac.uk/wp-content/uploads/2020/06/Caring-and-COVID-19_Hunger-and-mental-wellbeing-2.pdf](http://circle.group.shef.ac.uk/wp-content/uploads/2020/06/Caring-and-COVID-19_Hunger-and-mental-wellbeing-2.pdf)

[8] Bennett MR, Zhang Y, Yeandle S. Caring and COVID-19 - Loneliness and use of services. Care Matters Series Published Online First: 2020.[http://circle.group.shef.ac.uk/wp-content/uploads/2020/08/CARING-and-COVID-19-Loneliness-and-use-of-services_04.08.20.pdf](http://circle.group.shef.ac.uk/wp-content/uploads/2020/08/CARING-and-COVID-19-Loneliness-and-use-of-services_04.08.20.pdf)

[9] Lorenz-Dant K, Comas-Herrera A. Pre-print: The impacts of COVID-19 on unpaid carers of adults with long-term care needs and measures to address these impacts: a rapid review of the available evidence (2021). [https://ltccovid.org/2021/01/15/pre-print-the-impacts-of-covid-19-on-unpaid-carers-of-adults-with-long-term-care-needs-and-measures-to-address-these-impacts-a-rapid-review-of-the-available-evidence/](https://ltccovid.org/2021/01/15/pre-print-the-impacts-of-covid-19-on-unpaid-carers-of-adults-with-long-term-care-needs-and-measures-to-address-these-impacts-a-rapid-review-of-the-available-evidence/)

[10] Gallagher, Stephen, and Mark A. Wetherell. 2020. Risk of Depression in Family Caregivers: Unintended Consequence of COVID-19. BJPsych Open 6 (6). Royal College of Psychiatrists. (https://doi:10.1192/bjo.2020.99)[https://doi:10.1192/bjo.2020.99]

[11]Allen G: We must not forget those everyday heroes, the unpaid carers. Future Care Capital Published Online First: 2021 [https://futurecarecapital.org.uk/latest/we-must-not-forget-those-everyday-heroes-the-unpaid-carers/](https://futurecarecapital.org.uk/latest/we-must-not-forget-those-everyday-heroes-the-unpaid-carers/)

## Section 2 - Who is providing unpaid caring? {.tabset}

### Change in unpaid caring during the pandemic

This analysis explores the change in caring status during the pandemic. 

```{r message = FALSE, fig.align="centre"}
##Sankey Diagram 
df<-all %>% 
  select(carer_pre,care_hours_pre,care_loc_cv,care_loc_change,care_hours,betaindin_lw, betaindin_xw, psu, strata) %>% 
  filter(!is.na(betaindin_lw))

sankey_flow_addNUM<- df %>% 
  mutate(flow_1=case_when(care_hours_pre != "" ~ paste0(care_hours_pre, sep='_1')),
         flow_2=case_when(care_hours != "" ~ paste0(care_hours, sep='_2'))) %>% 
  select(flow_1, flow_2,betaindin_lw,psu,strata) %>% 
  drop_na()

uos_design<-svydesign(id= ~psu, strata= ~strata, survey.lonely.psu="adjust",
                      weights= ~betaindin_lw, data=sankey_flow_addNUM)


sankey_plot_data<-svytable(~flow_1+flow_2, design=uos_design) %>% 
  as.data.frame() %>%
  rename(In=flow_1, Out=flow_2) 
  

nodes<- sankey_plot_data %>% 
  select(In, Out) %>% 
  pivot_longer(c(In,Out), names_to="col_names", values_to= "name_match") %>% 
  select(-1) %>% 
  distinct() %>% 
  mutate(name=str_sub(name_match, end=-3))

nodes<-data.frame(nodes)

sankey_plot_id<- sankey_plot_data %>% 
  mutate(IDin= match(In, nodes$name_match)-1,
         IDout=match(Out, nodes$name_match)-1)

sankey_plot_id<-data.frame(sankey_plot_id)

# node_colour<-'d3.scaleOrdinal() .domain(["THF_red","THF_50pct_light_blue","THF_1_purple"])
#               .range(["#dd0031","#53a9cd","#744284"])'

# nodes<- nodes %>% 
#   mutate(ColourGroup=case_when(name=="No caring" ~ "THF_red",
#                                name== "Low Level Caring" ~ "THF_50pct_light_blue",
#                                name== "High Level Caring" ~ "THF_1_purple"))
# 
# sankey<-sankeyNetwork(Links = sankey_plot_id, Nodes = nodes, 
#                       Source = "IDin", Target = "IDout", 
#                       Value = "Freq", NodeID = "name", sinksRight = FALSE, fontSize = 14, colourScale = node_colour,  NodeGroup = "ColourGroup", width=1000, height=800)
# 
# sankey

nodes<- nodes %>% 
  mutate(ColourGroup=case_when(name=="No caring" ~ "#dd0031",
                               name== "Low Level Caring" ~ "#53a9cd",
                               name== "High Level Caring" ~ "#744284"))

fig <- plot_ly(
  type = "sankey",
  orientation = "h",
  
  node = list(
    label = nodes$name,
    color = nodes$ColourGroup,
    pad = 15,
    thickness = 20,
    line = list(
      color = "black",
      width = 0.5
    )
  ),
  
  link = list(
    source = sankey_plot_id$IDin,
    target = sankey_plot_id$IDout,
    value =  sankey_plot_id$Freq
  )
)
fig <- fig %>% layout(
  title = "Change in unpaid caring during pandemic",
  font = list(
    size = 12
  )
)

fig


```

**Summary:**

* 77% of those providing low level care and 43% providing high level care during the second half of the pandemic did not identify as carers pre pandemic. New unpaid carers and those who have started providing more intensive care might not yet have access to information and support for carers.

### Sandwich Carers

In this analysis, sandwich carers are defined as people aged 16 to 70 who have at least one child in their household under 16 years old and are also providing unpaid care.This is similar to the definition used by the [ONS on their analysis of unpaid carers](https://www.ons.gov.uk/peoplepopulationandcommunity/healthandsocialcare/healthandwellbeing/datasets/sandwichcarers) 

```{r warning=FALSE, fig.align= "centre"}
sandwich_care<- all_dem %>% 
  filter(care_hours!= "No caring") %>% 
  select(carer, care_hours,sex_lab,age,hh_child_u16,betaindin_lw, betaindin_xw, psu, strata) %>% 
  mutate(sandwich=factor(case_when(hh_child_u16 =="Atleast One" & age < 71 ~ 1,
                                   (hh_child_u16== "None" | age >= 71) ~ 2),levels=c(1,2), labels=c("Sandwich carers","Not Sandwich carers")),
         care_hours=factor(care_hours, levels=c("Low Level Caring", "High Level Caring"), labels=c("Low Level Caring", "High Level Caring"))) %>% 
  drop_na()

uos_design<-svydesign(id= ~psu, strata= ~strata, survey.lonely.psu="adjust",
                      weights= ~betaindin_xw, data=sandwich_care)


tab_s<-svytable(~sex_lab+ sandwich +care_hours, design=uos_design) %>% 
  as.data.frame() %>% 
  group_by(care_hours) %>% 
  mutate(sum_care=sum(Freq)) %>% 
  ungroup() %>% 
  mutate(prop_all=Freq/sum_care) %>% 
  unite("Metric",c(sex_lab, sandwich), sep= " ", remove=FALSE)


g<-tab_s%>% 
  ggplot(., mapping=aes(x=care_hours, y=prop_all*100, label=percent(prop_all), fill=Metric))+
  geom_bar(position="stack", stat="identity")+
  geom_bar_text(position = "stack", reflow = TRUE)+
  scale_x_discrete(labels = label_wrap(20))+
  scale_y_continuous(limits = c(0,100), breaks = seq(0, 100, by = 10))+
  scale_fill_THF()+
  theme_THF()+
  labs(x= "", y="", title="Sandwich Carers by Sex and Type of Caring")+
  theme(plot.title = element_text(size=10),
        legend.text=element_text(size=9), legend.position = "top", axis.text.x=element_text(size=9),axis.text.y=element_text(size=9))

ggsave(here::here(here::here('outputs', 'sandwich_care.png')),dpi=300,
         width = 10, height = 6.5) 


ggplotly(g, tooltip=c("x","label"))




```

**Summary:**

* Regardless of type of caring, Women account for a higher proportion of carers during the second wave of the pandemic.
* However, women with children in the household are over-represented in the “high level of care” group. More than 1 in 5 (20%) carers providing high level caring are women with children in the household, this is compared to 1 in 10 among carers providing low level caring.

## Implications

* Carers do not always self-identify and new carers especially might not yet have access to information and support for carers. We need to re-double efforts to identify and support unpaid carers, as signalled in NHS Long Term plan, and to develop a National Carers strategy.  
* Given the increase in high level care i) need to ensure there is adequate respite care available, ii) need to support flexible working so that those who want to combine work and care can do so, iii) need to provide financial support for those who can't combine work and caring.


## Section 3: What is the state of health of our unpaid carers? {.tabset}

### Mental Health

Previous studies have reported that carers during the pandemic have a higher risk of depressive symptoms.[11] Using the same methodology to measure mental health, this analysis is reports on carers' mental health during the second half of the pandemic. 
```{r, message=FALSE, fig.align= "centre"}


GHQ<- all_h %>% 
  select(carer, care_hours, GHQ_cv, betaindin_xw, psu, strata) %>% 
  mutate(GHQ_label=case_when(GHQ_cv=="Non-DS"~ "No Depressive Symptoms",
                             GHQ_cv=="DS"~ "Depressive Symptoms"))


uos_design<-svydesign(id= ~psu, strata= ~strata, survey.lonely.psu="adjust",
                      weights= ~betaindin_xw, data=GHQ)


tab_s_dy<-svytable(~GHQ_label+care_hours, design=uos_design) %>% 
  as.data.frame() %>% 
  group_by(care_hours) %>% 
  mutate(sum_care=sum(Freq)) %>% 
  ungroup() %>% 
  mutate(prop_all=Freq/sum_care)


g2<-tab_s_dy%>% 
  ggplot(., mapping=aes(x=care_hours, y=prop_all*100, label=percent(prop_all), fill=GHQ_label))+
  geom_bar(position="stack", stat="identity")+
  geom_bar_text(position = "stack", reflow = TRUE)+
  scale_x_discrete(labels = label_wrap(10))+
  scale_y_continuous(limits = c(0,100), breaks = seq(0, 100, by = 10))+
  # facet_grid(cols=vars(care_hours))+
  scale_fill_THF()+
  theme_THF()+
  labs(x= "", y="", title="Mental Health during pandemic by caring type")+
  theme(plot.title = element_text(size=10),
        legend.text=element_text(size=10), legend.position="top", axis.text.x=element_text(size=10),axis.text.y=element_text(size=10))
 

ggsave(here::here(here::here('outputs', 'Mental_Health.png')),dpi=300,
         width = 10, height = 6.5) 


ggplotly(g2, tooltip=c("x","label"))

 
```

**Summary**

* High level carers have the poorest mental health during the pandemic (compared with low level carers and those not providing unpaid care). 

### Number of Long Term Health conditions

```{r, message=FALSE, fig.align= "centre"}


MLTC<- all_h %>% 
  select(care_hours,mltc, betaindin_xw, psu, strata) %>% 
  filter(!is.na(betaindin_xw)) %>% 
  mutate(mltc_short=case_when(mltc %in% c("None", "One")~ "None/One",
                              mltc=="2+"~ "2 or more"))


uos_design2<-svydesign(id= ~psu, strata= ~strata, survey.lonely.psu="adjust",
                      weights= ~betaindin_xw, data=MLTC)


tab_s<-svytable(~mltc_short+ care_hours, design=uos_design2) %>% 
  as.data.frame() %>% 
  group_by(care_hours) %>% 
  mutate(sum_care=sum(Freq)) %>% 
  ungroup() %>% 
  mutate(prop_all=Freq/sum_care)



g3<-tab_s%>% 
  ggplot(., mapping=aes(x=care_hours, y=prop_all*100, label=percent(prop_all,0.01), fill=mltc_short))+
  geom_bar(position="stack", stat="identity")+
  geom_bar_text(position = "stack", reflow = TRUE)+
  scale_x_discrete(labels = label_wrap(10))+
  scale_y_continuous(limits = c(0,100), breaks = seq(0, 100, by = 10))+
  # facet_grid(cols=vars(care_hours))+
  scale_fill_THF()+
  theme_THF()+
  labs(x= "", y="", title="Number of health conditions by caring type")+
  theme(plot.title = element_text(size=10),
        legend.text=element_text(size=10), legend.position="top", axis.text.x=element_text(size=10),axis.text.y=element_text(size=10))

ggsave(here::here(here::here('outputs', 'mltc.png')),dpi=300,
         width = 10, height = 6.5) 


ggplotly(g3, tooltip=c("x","label"))


```


**Summary**

* Unpaid carers, particularly those in the high level caring group have more long term conditions than the no caring group.

## Implications

* Signpost carers to mental health and wellbeing support. We also need to support carers with their financial and social needs because these financial hardship and low social connectedness are strongly linked to poor mental health.


## Section 4: How have our unpaid carers accessed healthcare services during the pandemic? {.tabset}

### Access to services

```{r message=FALSE, fig.align= "centre"}

all<-all_s %>% 
  left_join(all_dem)

services<- all %>% 
  select(care_hours_pre,care_hours, carer, carer_pre, wait_for_NHS_treat, respite, betaindin_lw,betaindin_xw, psu, strata) %>%
  mutate(new_care= factor(case_when(carer=="Yes"& carer_pre=="Yes"~0,
                                    carer=="Yes"& carer_pre=="No"~1,
                                    carer=="No"& carer_pre=="Yes"~1,
                                    carer=="No"& carer_pre=="No"~0), levels=c(0,1)))


uos_design3<-svydesign(id= ~psu, strata= ~strata, survey.lonely.psu="adjust",
                      weights= ~betaindin_xw, data=services)

tab_s_dy<-svytable(~wait_for_NHS_treat+care_hours, design=uos_design3) %>% 
  as.data.frame() %>% 
  group_by(care_hours) %>% 
  mutate(sum_care=sum(Freq)) %>% 
  ungroup() %>% 
  mutate(prop_all=Freq/sum_care)


g4<-tab_s_dy%>% 
  ggplot(., mapping=aes(x=care_hours, y=prop_all*100, label=percent(prop_all), fill=wait_for_NHS_treat))+
  geom_bar(position="stack", stat="identity")+
  geom_bar_text(position = "stack", reflow = TRUE)+
  scale_x_discrete(labels = label_wrap(10))+
  scale_y_continuous(limits = c(0,100), breaks = seq(0, 100, by = 10))+
  # facet_grid(cols=vars(care_hours))+
  scale_fill_THF()+
  theme_THF()+
  labs(x= "", y="", title="Since 1st Jan 2020, have you been waiting for NHS treatment?")+
  theme(plot.title = element_text(size=10),
        legend.text=element_text(size=10), legend.position="top", axis.text.x=element_text(size=10),axis.text.y=element_text(size=10))

ggsave(here::here(here::here('outputs', 'access.png')),dpi=300,
         width = 10, height = 6.5) 

ggplotly(g4, tooltip=c("x","label"))

```

**Summary**

* Many unpaid carers have experienced cancelled or postponed NHS treatment. While this doesn’t look different for unpaid carers vs others, there’s an argument to prioritise their care given that their responsibilities. 

## Implications

* We need to avoid the longer term effects of COVID-19 on carer health through missed health care. Prioritise unpaid carers for NHS treatment during the recovery. This is important because high level carers have more LTCs and because of the consequences of their ill health on the people they care for. This is going to need providers to better identify unpaid carers. Echo the LT plan calls to improve how we identify and support carers.








